SUPPLEMENTARY INFORMATION: NMRlipids Databank: Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

### 2022-08-31 SI.tex

\title{NMRlipids Databank: Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration}

@@

\begin{abstract}
We present a databank of lipid bilayer simulations from the NMRlipids open collaboration project.
\end{abstract}
\begin{document}

@@

\section{Introduction}

%The demand for sharing and reusing of MD simulation data is increasing, but practical solution remains unclear due to unsolved issues in data storage and indexing.

The importance of sharing MD simulation data following the FAIR principles~\cite{wilkinson16} has been widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20}, and databanks are emerging~\cite{meyer10,kamp10,hospital16,mixcoha16,newport19,bekker20,espigares20,leston22}.
The relevance of quality evaluation of simulation trajectories in databanks regarding technical details of simulations and accuracy of the underlying physical description of the system (force field) has become evident~\cite{tai04,meyer10,hospital20} and such quality evaluation has in some cases also been implemented~\cite{meyer10,hospital16}. However, straightforward quality comparisons between individual simulations or force fields within these databanks remain challenging. 
While importance of such databanks for MD simulations is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and different kinds of approaches are emerging~\cite{meyer10,kamp10,hospital16,mixcoha16,newport19,bekker20,espigares20,leston22}, generally accepted protocols and best practices are still under active development.

Here we present a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  The concept of overaly databank is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it pontentially used for wide range of situations, particularly when storage of raw data requires significant resources and final outcomes or best practices are not yet clear, overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.

The practical relevance of the NMRlipids databank is exemplified by automatic quality evaluation and ranking of large amount of MD simulation data, data-driven analysis detecting correlations between properties of model cell membranes and analyses of rare phenomena that are beyond the scope of standard MD simulation studies. The NMRlipids databank provides new tools for researchers in wide range of fields in academia and industry from cell membrane biology to lipid nanoparticle formulations and data-driven computational chemistry and machine learning.

%The Introduction section, of referenced text\cite{Figueredo:2009dg} expands on the background of the work (some overlap with the Abstract is acceptable). The introduction should not include subheadings.

\section{Results}

%Up to three levels of \textbf{subheading} are permitted. Subheadings should not be numbered.
\subsection{Design of the NMRlipids databank}
The key idea of the overlay databank is that the storage of raw data in layer 1 is distributed in publicly available repositories or other servers with long term stability and permanent links such as digital object identifiers. The core of the databank, layer 2, contains only information on the location and content of the raw data, thereby not requiring large resources to handle and maintain. This lowers the barrier for starting such databank as well as for long term storage. The NMRlipids databank is essentially a git repository containing information on the location of raw data and its indexing with universal naming convention. In addition to all computers where the databank is developed and used, the NMRlipids databank git is stored to Zenodo server, thereby enabling a very cost effective long term storage for the databank. The databank can be used in layer 3 by accessing the raw data and information stored in the databank by employing the universal naming conventions for atoms, molecules and simulation details. The applications can be linked to the core databank (layer 2) without actually including them, thereby enabling flexible extension of the data without compromising the simplicity and lightness of the core databank.

\begin{figure}
    \centering
    \includegraphics[width = 180mm]{Figures/overlay.pdf}
    \caption{a) Structure of an overlay databank. 
    %Raw data is available from a publicly accessible repository (layer 1).
    %The core of the databank contains information on the location of the raw data and indexing of molecular %compositions and simulation details described with universal naming conventions (layer 2).
    %The content of databank can be viewed and analysed using external programs that automatically read the information and data from layers 1 and 2 (layer 3). Examples of such applications are the GitHub repository generating the results presented in this work and the graphical access to the databank.
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    The figures and numbers are created on 9th of May 2022.}
    \label{fig:overlay}
\end{figure}

Currently the databank is composed of approximately 500 trajectories with the total length of approximately 231 microseconds of which most are contributed for the previous publications from the NMRlipids open collaboration~\cite{botan15,catte16,antila19,bacle21}. The distribution of lipids, force fields, length of the trjajectories and available binary mixtures are shown in Fig.~\ref{fig:overlay}.

\subsection{Quality evaluation of force fields}
Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19}, but universal quality measure for membrane simulations is not defined and controversial results are often reported from simulations~\cite{antila22b}. The unclarities in simulation quality complicate the selection of proper simulation models for specific applications and estimation of reliability of reported simulation results, thereby being a major obstacle in many applications of membrane MD simulations.

To enable rapid evaluation and comparison of membrane MD simulation qualities, we define the quality measure for membrane MD simulations using the C-H bond order parameters from NMR experiments and form factors from x-ray scattering, which are robust experimental measurables that can be directly connected to the simulation data~\cite{ollila16}. C-H bond order parameters are related to the conformational ensemble of individual lipid molecules, while the form factor to arises from electron density thereby connecting to the overall structure of lipid bilayers. This is demonstrated in Fig.~\ref{fig:quality} a) showing correlations between membrane lateral density (area per lipid), thickness, minima of form factor and acyl chain order analyzed from all apporimately 500 trajectories in the NMRlipids databank. Membrane area per lipid has negative correlations with membrane order and thickness with Pearson coefficient of -0.78 and -0.49, respectively, because decreased area leads to more ordered, tightly packed and stretched acyl chains leading to thicker membranes. On the other hand, locations of form factor minima have positive correlation with the membrane area and negative correlation with the thickness. In conclusion, both x-ray scattering form factors and acyl chain order parameters from NMR are on average good proxies for membrane packing and thickness, while NMR order parameters give segmental resolution information on the quality of conformational ensembles of lipids.

\begin{figure}[tbp]
    \centering
    \includegraphics[width=180mm]{Figures/quality.pdf}
    \caption{a) Correlations between membrane area per lipid, thickness, second minima of form factor and average order parameter of the {\it sn}-1 acyl chain extracted from the NMRlipids databank. All Pearson correlation coefficients shown in bottom left corner have p-value below 0.001.
    b) Evaluation against experimental data exemplified for the simulation giving the best qualities in overall, for headgroup and x-ray scattering form factor.
    c) Change in area per lipid upon addition of POPE to POPC in different simulation extracted from the NMRlipids databank and quality evaluation table of POPE simulations.
    d) Quality evaluation table showing the best 13 simulations according to the overall order parameter quality.}
    \label{fig:quality}
\end{figure}

NMRlipids databank contains also experimental order parameter and x-ray form factor data that is connected to corresponding  simulations in order to define the simulation qualities and rank them to select the suitable simulation models for particular applications. The ranking of all simulations based on estimated average probability of order parameters to locate within experimental errors are shown in Fig.~\ref{fig:quality} b) and the comparison for the best models for overall quality, headgroup and glycerol backbone, and form factor are exemplified in Fig.~\ref{fig:quality} d). Full details of the NMRlipids quality measure for lipid bilayer simulations are given in the methods section and supplementary information.
The top ranking simulations in Figs.~\ref{fig:quality} b) and d) demonstrate also the current complexity in lipid bilayer simulation quality. Simulation of POPC lipid bilayer with Slipids force field ranks first in overall score, glycerol backbone is not correcetly captured in this model. On the othere, simulation with the best quality for headroup and glycerol backbone, CHARMM36 POPC, predicts too ordered acyl chain, thereby not being within the top 10 simulations in the overall ranking. The best ranking for x-ray form factor is given by the Slipids POPE simulation, however, because the quantitative value for form factor quality depends on the quality of experimental data, form factor qualities can be compared only between simulations evaluated against the same experimental dataset.

The power of NMRlipids quality metrics in selecting the best model for particular application is demontrated in Fig.~\ref{fig:quality} c) where the area per lipids of POPC:POPE mixtures from different force fields are shown and quality of POPE simulation in different force fields are evaluated. Among the simulations available in the NMRlipids databank, the Slipids force field gives the best quality in terms of acyl chain order parameters and form factor, and predicts the largest area per molecule for POPE and smallest difference with POPC. In conclusion, simulations with Slipids is the most reliable force field for membrane packing in POPC:POPE mixtures althought its glycerol backbone is not accurately modelled. Similar comparisons utilizing the preliminary data from the NMRlipids databank have concluded that Slipids is relatively good model also for mixtures of charged POPC:POPS membranes, although models with better counterion binding have higher quality, but CHARMM36 is the best to study differences in headgroup conformational ensembles between different lipid types. Understanding such complex picture of lipid bilayer MD simulation quality would not be possible without automatic quality evaluation of large sets of simulations enabled by the NMRlipids databank.

\subsection{Water diffusion anisotropy in membrane systems}
The anisotropic water diffusion through and along of membranes is utilized in MRI imaging \cite{??} and plays a role in the drug translocation through biological material, particularly in skin \cite{hansen13,wen18,nitsche19,roberts21,??}. Dependencies between membrane properties, water permeablity, and anisotropic water diffusion have been recognized \cite{??} but consensus on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, has not yet been reached despite of several available models~\cite{nitsche13,nitsche16,shinoda16,venable19,??}. MD simulations can be used to elucidate such dependencies, but systematic collection of sufficient amount of data has been challenging because few water permeation events are typically observed in a single MD simulation trajectory \cite{??}. Nevertheless, averaging over large amounts of data available in the NMRlipids databank enables to study the trends of water permeation upon changing memebrane properties.

%averaging Here we demonstrate the power of the NMRlipids databank by analysing how 
Water permeabilities calculated from all the simulations in the NMRlipids databank as a function of area per lipids, thickness, and acyl chain order are shown in  Fig.~\ref{fig:permeability} (a-c). To illustrate how permeability depends these properties, average over systems with fixed values of area per lipids, thickness, and acyl chain order are also shown in the figure. Permeability increases with increasing area per lipid and decreases with increasing membrane thickness and acyl chain order. Previous simulation and experimental studies have reported linear dependence of permeability on both area per lipid and thickness for limited set of lipid mixtures but not always for all types of lipid mixtures \cite{mathai08,frallicciardi22}. Our results suggest stronger dependence for thicknesses below 3.9\,nm and areas above 69\,Å$^2$, but the dependence may appear linear above and below these values as shown in the insets of Figs. \label{fig:permeability} a) and b). As expected, the permeability increases with increasing temperature in Figs. \label{fig:permeability} d). The Arrhenius plot gives 17\,$\pm$\,4\,k$_b$T for the average energy barrier for the water permeation. The values for water permeabilities from simulations in the databank vary between 0.3 and 322\,$\mu$m/s with the mean and median of 14\,$\mu$m/s and 8\,$\mu$m/s, respectively. These values are within the range of experimentally reported values and the mean is the same order of magnitude as the diffusive permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. Clear dependencies of permeability on hydration level, charged lipid fraction, cholesterol fraction or POPE fraction were not observed, although weak decrease for the latter two may be visible.

\begin{figure}[tbp]
    \centering
    \includegraphics[width=180mm]{Figures/permeation2.pdf}
    \caption{Water permeation through lipid membranes analyzed from the databank as a function of 
      a) area per lipid,
      b) thickness,
      and c) temperature. 
      Inserts in a) and b) show the region where the dependence could be considered approximately linear.
      Insert in c) shows the Arrhenius plot of permeation ($\ln$ (P) vs. 1/T) that gives 17\,$\pm$\,4\,k$_b$T for the average activation energy for water permeation through lipid bilayer.}
    \label{fig:permeability}
\end{figure}

To analyze the anisotropic water diffusion, we also calculated the water diffusion along the membrane. The results at different hydration levels are shown in Fig.~\ref{fig:permeability} e) together with the experimental data~\cite{??}. The experimental water diffusion increases with increasing hydration level toward the value for bulk water (?? in ??K). Simulations overestimate the experimental bulk value approximately with the factor of 2 at high hydration levels which is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion. However, simulation results are closer to experiments with low hydration levels. To estimate the diffusion anistoropy, we translated the water permation to diffusion coefficent through multilamellar stack using the Tanner equation~\cite{??}. The results in

\section{Discussion}

%The Discussion should be succinct and must not contain subheadings.

Quality measure and automatic quality evaluation of lipid bilayer MD simulations introduced in the NMRlipids databank enables rapid ranking of available simulation models agains experimental NMR and x-ray scattering data. This provides a tool for researches to rapidly evaluate the credibility of MD simulations of their own and reported by other groups. Because such tool and quality measure has not been available, this will elaborate the quality of published MD simulations of lipid bilyers and reduce potentially misleading results~\cite{antila22b}. The power of NMRlipids databank to select the best models for particular applications has been demontrated for PC/PE lipid mixtures (Fig.~\ref{fig:quality}), PC/PS lipid mixtures~\cite{antila22b}, and lipid headgroups~\cite{bacle21}.

The increasing amount of MD simulation data with programmatic access in the NMRlipids databank opens up possibilities for wide range of applications utilizing the large set of accessible data. Extend of the data in the NMRlipids databank in terms of quantity (e.g., simulation length and number of conformations), content (e.g, lipid compositions and ion concentrations) and quality enables analyses that are not possible to conduct from MD simulation data produced by a single research group. Applications of the NMRlipids databank to understand how diffusion of water through and along cellular membrane depends on its physical properties are demonstrated in Fig.~\ref{??}. Permeation of water through membranes resembles the permeation of also other hydrophilic molecules, such as drugs, and detailed understanding of water dynamics through and along membranes is potentially useful for the development of MRI imaging methods~\cite{topgaard20}. 
These examples demonstrate the practical applications of NMRlipids databank on problems in the biological and biomedical sciences.

Building accessible databanks of molecular dynamics simulation data has been challenging due to the required long term support for hardware and software maintenance. In the overlay model used in the NMRlipids databank, the demand of hardware can be distributed and open collaboration model reduces the risk for ending software maintenance. Furthermore, the open collaboration model used in the NMRlipids project credits contributors by offering authorship in published articles, thereby creating an incentive for contributions. This model could be extended also to other fields where similar barriers to establish publicly accessible databanks exist. Emerging applications of machine learning are increasing the impact of such databanks. For example, the existing Protein Databank (PDB)~\cite{??} containing experimentally determined protein structures with programmatic access has enabled the development of machine learning based tools building on the data collected in the databank over the years~\cite{??}. NMRlipids and other databanks with open programmatic access have potential to lead similar unforeseen applications in the future.

\section{Methods}

%Topical subheadings are allowed. Authors must ensure that their Methods section includes adequate experimental and characterization data necessary for others in the field to reproduce their work.

\subsection{Structure of the databank}

\begin{figure}
  \includegraphics[width=\textwidth]{Figures/DataBankChart.pdf}
  \caption{Structure of the NMRlipids databank. Manually added input data (blue boxes) includes basic information on the simulation, permanent links to the raw data, and experimental data if available. The databank entries (red box) and analysis results (green boxes), locating at \url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations} are automatically generated by the computer programs included in the NMRlipids databank (yellow boxes). Because raw data are not permanently stored but can be accessed based on the information in the databank, this connection is marked with the dashed line. }\label{DatabankStructure}
\end{figure}

Structure of the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure}. The required input information to create an entry into the NMRlipids databank are listed in table~\ref{tab:READMEkeys}. While the raw simulation data is not directly stored in the NMRlipids databank, permanent links from where the raw data can be accessed have to be given and are then stored in the README.yaml files at \url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations}. These files contain all the essential information listed in table~\ref{tab:READMEkeys} on each simulation entry that are needed for further applications. The raw MD simulation data can locate in any stable publicly available repository, although all the current data locates in Zenodo \url{www.zenodo.org}.

%A README.yaml file is made using AddData.py script for each simulation stored in the databank based on the information given by the author of the simulation data. In addition to this information, automatically extracted information about the simulation are stored in the README.yaml files (see table~\ref{tab:READMEkeys}). A README.yaml file for each simulation is stored to a folder named according to the hashes of original trajectory and topology file locating in \url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations}. These README.yaml files give access to all the essential information on the simulations in the databank either directly, or pointing to a original file from where the information can be extracted.

\begin{table}[]
    \centering
    \begin{tabular}{  p{3.5cm}  p{9.5cm}  p{4.0cm} }
    \toprule
    key & description & type  \\
    \midrule
        %\hline
    DOI & DOI from where the raw data is found & user given (compulsory) \\
    SOFTWARE & Software used to run the simulation (e.g. Gromacs, Amber, NAMD, etc.) & \\
    TRJ & Name of the trajectory file found from DOI & \\
    TPR & Name of the topology file found from DOI (trp file in the case of Gromacs) & \\
    PREEQTIME & Pre-equilibrate time simulated before the uploaded trajectory in nanoseconds. \tablefootnote{For example, if you upload 100-200 ns part of total 200 ns simulation, this should value should be 100.} & \\
    TIMELEFTOUT & Equilibration period in the uploaded trajectory that should be discarded in analyses. \tablefootnote{For example, if you upload 0-200 ns part of total 200 ns simulation where the first 100 ns should be considered as an equilibration, this value should be 100.} \\
    COMPOSITION & Molecules names used in the simulation and corresponding mapping files (see section \ref{naming}) & \\
    DIR\_WRK & Temporary working directory in your local computer. \\
    UNITEDATOM\_DICT & Information for constucting hydrogens for united atom simulations, empty for all atom simulations & \\
    TYPEOFSYSTEM & Lipid bilayer or something else & \\
    \hline
    PUBLICATION & Give reference to a publication(s) related to the data. & User given (optional)\\
    AUTHORS\_CONTACT & Name and email of the main author(s) of the data. & \\
    SYSTEM & System description on free text format & \\
    SOFTWARE\_VERSION & Version of the used software & \\
    FF & Name of the used force field & \\
    FF\_SOURCE & Source of the force field parameters, e.g, CHARMM-GUI, webpage, citation to a publication, etc. & \\
    FF\_DATE &  Date when force field parameters were accessed on the gives source (day/month/year). & \\
    FF{molename} & Molecule specific force field information, e.g., water model with FFSOL and sodium parameters with FFSOD. & \\
    CPT & Name of the Gromacs checkpoint file. & \\
    LOG & Name of the Gromacs log file. & \\
    TOP & Name of the Gromacs top file. & \\
    GRO & Name of the Gromacs gro file. & \\
    \hline
    TRAJECTORY\_SIZE & Size of the trajectory file in bytes & automatically extracted data. \\
    TRJLENGTH & Lenght of the trajectory (ps). & \\
    TEMPERATURE & Temperature of the simulation. & \\
    NUMBER\_OF\_ATOMS & Number of atoms in the simulation. & \\
    DATEOFRUNNIG & Date when added into the databank & \\
    EXPERIMENT & Potentially connected experimental data & \\
    COMPOSITION & Numbers of lipid molecules (NPOPC, NPOPG, etc.) per membrane leaflet are calculated by determining on which side of the center of mass of the membrane the center of mass of the head group of each lipid molecule is located.
    Numbers of other molecules such as solvent and ions (NSOL, NPOT, NSOD, etc.) are read from the topology file. & \\
    \end{tabular}
    \caption{Keys stored in the README.yaml files of simulations.}
    \label{tab:READMEkeys}
\end{table}

In order to evaluate the quality of simulations, sets of C-H bond order parameters from NMR and from factors from x-ray scattering are included in the NMRlipids databank (\url{https://github.com/NMRLipids/Databank/tree/main/Data/experiments}). The required information for an experimental dataset are listed in table~\ref{tab:READMEkeysEXP}.
A simulation is connected to a experimental data set
%C-H bond order parameter and x-ray scattering form factor data using the searchDATABANK.py script. Available experimental data is collected to  where each dataset is saved into a folder with the name based on DOI of the related publication and experimental conditions are described in README.yaml files with the keys listed in table~\ref{tab:READMEkeysEXP}. 
if molar concentrations of all molecules are within $\pm$5 percentage units, charged lipids have the same counterions, and temperature is within $\pm$2 degrees. In such cases, a simulation and experimental data are paired by adding the path to the experimental data into the simulation README.yaml file. 
%If lipid to water ratio is above ??, the systems are considered fully hydrated. Otherwise also hydration level is considered.

\begin{table}[]
    \centering
    \begin{tabular}{  p{5.0cm}  p{10.0cm}}
    \toprule
    key & description \\
    \midrule
    DOI & DOI of the publication related to the experimental data. \\
    TEMPERATURE & Temperature of the experiment. \\
    MOLAR\_FRACTIONS & Dictionary of molar fractions of bilayer components \\
    ION\_CONCENTRATIONS & Dictionary of ion concentrations of the system \\% (defined as ??) \\
    TOTAL\_LIPID\_CONCENTRATION & Total concentration of lipid components. If exact concentration is not known, but experiments are performed in excess water, 'full hydration' can be given. \\
    COUNTER\_IONS & Type of counter ions if present.
\end{tabular}
    \caption{Keys stored in the README.yaml files of experiments.}
    \label{tab:READMEkeysEXP}
\end{table}

Because README.yaml contains all the essential information on each simulation, arbitrary analyses can be automatically performed over all the simualtions in the NMRlipids databank. In addition to the order parameters for each C-bonds and x-ray scattering form factors used in the quality evaluation, the NMRlipids databank contains the area per lipid and thickness calculated from all simulations in the databank. These results are stored in the same folders as the README.yaml files in \url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations}.

\subsection{Molecule and atom naming convention} \label{naming}

Because universal convention for lipid molecules and atoms therein has not been defined, the naming conventions vary between authors and force fields. To enable automatic analyses over large sets of simulation in the NMRlipids databank, we have defined unique naming conventions for lipid molecules and atoms. The abbreviations of molecule names used in the NMRlipids databank are listed in table~\ref{tab:abbreviations}. The unique atom names for each molecule and corresponding names in each simulation are defined using mapping files introduced in the NMRlipids project (\url{https://nmrlipids.blogspot.com/2022/04/new-yaml-format-of-mapping-files.html}).

Molecule and atom names in each simulation are connected to the unique naming convention with the COMPOSITION dictionary in the README.yaml file. Upon addition of a new entry in the databank, the molecule names in the simulation corresponding the unique names and mapping files (available at \url{https://github.com/NMRLipids/Databank/tree/main/Scripts/BuildDatabank/mapping_files}) are defined in the dictionary. The numbers of molecules in the system are then automatically calculated by the NMRlipids databank codes (AddData.py in Fig.~\ref{DatabankStructure}) and stored in the README.yaml together with other content of the COMPOSITION dictionary. This information can be then used to find each molecule and atom from each simulation in the analysis codes.

\begin{table}[h]
    \centering
    \begin{tabular}{c|c}
        Abbreviation & Molecule name \\
        \hline
        POPC &  1-palmitoyl-2-oleoyl-sn-glycero-3-phosphocholine\\
        POPG &  1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoglycerol \\
        POPS & 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-L-serine \\
        POPE & 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine \\
        CHOL & cholesterol \\
        DHMDMAB & dihexadecyldimethylammonium \\
        \hline
        POT & potassium ion \\
        SOD & sodium ion \\
        CLA & chloride ion \\
        CAL & calcium ion \\
        SOL & water \\
    \end{tabular}
    \caption{Abbreviations used in the databank}
    \label{tab:abbreviations}
\end{table}

\subsection{Quality evaluation}

The quality of conformational ensembles of individual lipid molecules in simulations are evaluated using the C-H bond order parameters which can be directly compared with robust experimental data~\cite{ollila16}.
%The C--H bond order parameters were calculated directly
%from the carbon and hydrogen positions using the 
%definition
The order parameters are defined as
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
%(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
and angular brackets denote the ensemble average, i.e. average over all sampled configurations of all lipids in a simulation.
%by estimating
%the average probability of C-H bond order parameters to agree with the values from NMR experiments. 
%As a first step, the quality of each order parameter is evaluated using the negative logarithm of probability to find the simulated order parameter within the experimental error bars
%\begin{equation}
%    S_q = -\log_{10}(P),    
%\end{equation}
Because conformational ensembles of individual lipids are independent in a simulation of a fluid lipid bilayer, the number of independent sample points for each C-H bond, $n$, equals the number of lipids in a simulation. Denoting the $S_{\rm CH}$ from Eq.~\ref{OP} as the sample mean and $s$ its variance calculated over individual lipids, 
%the probability the true mean of the order parameter is given by the %Student's t-distribution 
%Because each C-H bond order parameter is calculated as an average over individual lipids in a simulation
%\begin{equation}\label{studentT}
$\frac{S_{\rm CH}-\mu}{s/\sqrt{n}}$
%\end{equation}
%where $\bar{X}$ is the sample mean (order parameter calculated from simulations),
%$s$ is the sample variance (variance of order parameters in different lipids)
%and 
%has a 
has a Student's t-distribution with $n-1$ degrees of freedom, where $\mu$ is the real mean of the order parameter. 
%shifted by $S_{\rm CH}$ and scaled by $s/\sqrt{n}$.
Because lipid bilayer simulations contain at least dozens of lipids, the Student's t-distribution could be safely approximated with a normal distribution. However, when calculating the probability for a simulation to locate within experimental order parameters, the normal distribution gives values below the numerical accuracy of computers for simulations values far from experiments. To avoid such numerical instability, we use the first order Student's t-distribution.% for the equation \ref{studentT}.
Therefore, we estimate the probability for the simulation order parameter to locate within experimental error bars from equation
\begin{equation}\label{probability}
  P = f \left( \frac{S_{\rm CH} - (S_{\rm exp}+\Delta S_{\rm exp})}{s/\sqrt{n}} \right) - f \left( \frac{S_{\rm CH} - (S_{\rm exp}-\Delta S_{\rm exp})}{s/\sqrt{n}} \right),
\end{equation}
where f(t) is the first order Student's t-distribution.
% shifted to $\bar{X}$ and scaled with $s$.

%We estimate the probability for the simulation order parameter to locate within experimental
%error bars from equation
%\begin{equation}\label{probability}
%  P = T(S_{\rm exp}+\Delta S_{\rm exp}) - T(S_{\rm exp}-\Delta S_{\rm exp}),
%\end{equation}
%where T is the first order Student's t-distribution shifted to $\bar{X}$ and
%scaled with $s$. 
%where the probability is calculated from the normal distribution
%\begin{equation}\label{gaussian}
%    P = \int_{S_{\rm exp}-\Delta S_{\rm exp}}^{S_{\rm exp}+
%    \Delta S_{\rm exp}}  \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2} {\rm d}x.
%\end{equation}
%$S_{\rm exp}$ and $\Delta S_{\rm exp}$ are experimental order parameter and its error, and $\mu$ and $\sigma$ are the mean order parameter and its standard deviation from simulations. The quality measure $S_q$ approaches to zero when probability for agreement between simulation and experimental results approach one, and increases when simulated and experimental values diverge.
The error of $\Delta S_{\rm exp} = 0.02$ is currently assumed for all experimental order parameters~\cite{ollila16}. Because phospholipids sample their conformational ensemble within nanosecond timescale~\cite{ferreira15}, all simulations in the databank would be sufficiently long to sample the realistic conformational phase of individual lipids. However, some force fields exhibit too slow dynamics which leads to large error bars in order parameter values~\cite{antila21a}. Because large error bars widen the Student's t-distribution in Eq.~\ref{probability} thereby artificially increasing the probability to find the simulated value within experimental error bars, the order parameters with error bars larger than the experimental error 0.02 are not included in the quality evaluation.

The qualities of different fragments within each lipid type in a simulation, $P^{\rm frag}[{\rm lipid}]$, are then estimated by averaging the probabilities for individual C-H bond order parameters to locate within experimental errors according to Eq.~\ref{probability}, and dividing the average with the percentage of order parameters for which the quality is available within the fragment, $p_{\rm frag}[{\rm lipid}]$,
\begin{equation}
    P^{\rm frag}[{\rm lipid}] = \frac{\langle P[{\rm lipid}]\rangle_{\rm frag}}{p_{\rm frag}[{\rm lipid}]},
\end{equation}
 where frag can be {\it sn}-1, {\it sn}-2, headgroup or total (all order parameters within a molecule). The overall quality of different fragments in a simulation are then defined as a molar fraction weighted average over different lipid components
\begin{equation}
    P^{\rm frag} = \sum_{\rm lipid} \chi_{\rm lipid} P^{\rm frag}[{\rm lipid}],
\end{equation}
where $\chi_{\rm lipid}$ is the molar fraction of a lipid in the bilayer.

Qualities of form factors were evaluated using the same approach as in SIMtoEXP program~\cite{kucerka10}. Because experiments give form factors only in relative intensity scale, they were scaled to the simulation data using equation
\begin{equation}
    k_e = \frac{\sum_{i=1}^{N_q} \frac{|F_s(q_i)||F_e(q_i)|}{(\Delta F_e(q_i))^2}}{\sum_{i=1}^{N_q} \frac{|F_e(q_i)|^2}{(\Delta F_e(q_i))^2}}.
\end{equation}
The quality of each form factor was then calculated from the equation
\begin{equation}
    \chi^2 = \frac{\sqrt{\sum_{i=1}^{N_q}(|F_s(q_i)|-k_e|F_e(q_i)|)^2/(\Delta F_e(q_i))^2}}{\sqrt{N_q-1}},
\end{equation}
where $F_s$ and $F_e$ are form factors from a simulation and experiment, respectively, and summation goes over the experimentally available $N_q$ points.

\subsection{Analysing simulations in the NMRlipids databank}

The README.yaml files contain all the essential information to perform arbitrary analyses of simulations in the NMRlipids databank, i.e., the permanent location of the original data and naming convention for all atoms and molecules in each system. In practise, the analyse codes contains a loop over all README.yaml files (i.e., simulations in the NMRlipids databank) which first downloads the raw simulation to a local computer and then uses the information about the atom and molecule naming conventions in README.yaml and mapping files to perform the desired analyses. For example, the code that calculates all C-H bond order parameters of all systems is available at \url{https://github.com/NMRLipids/Databank/blob/main/Scripts/AnalyzeDatabank/calcOrderParameters.py} and minimal example of a analysis code is available at \url{https://github.com/NMRLipids/Databank/blob/main/Scripts/AnalyzeDatabank/template.ipynb}.

While the order parameters, form factors, area per lipid and thickness are stored within the NMRlipids databank (\url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations}), further analyses can be conventiently stored in separate repositories with the same folder structure based on hash identities of trajectory and topology files. For example, results from further analyses performed here are stored in folders at \url{https://github.com/NMRLipids/DataBankManuscript/tree/main/Data}. Such organization of the data enables further upcycling of the analyzed data as similarly to the original NMRlipids databank repository.

For the permeation of water through membranes, the number of permeation events in each trajectory was first calculated using the code by Camilo et al.~\cite{camilo2022}, available at~\url{https://github.com/crobertocamilo/MD-permeation}. The permeation was then calculated from equation $P=r/2c_w$, where $r$ is the rate per time and area, and $c_w$ is the concentration of water in bulk with the value of 33.3679\,(nm)$^{-3}$~\cite{venable19}.

%\subsection{Indexing the simulation data}

%\subsubsection*{AddData.py}
%AddData.py is a script that builds a database that contains a dictionary file and analysis data of each simulation. The dictionary file contains information about the simulation. The script also calculates order parameters of all CH bonds of the lipids in the simulation. To add a simulation it must be first uploaded to Zenodo (www.zenodo.org). The trajectory and topology files of the simulation are downloaded to the working directory from Zenodo but these are not saved into the database. To add a simulation to the database the user has to give some essential information about the simulation. This is done by writing a info file (*INFO.yaml) which is passed to AddData.py. 
%\newline \\
%AddData.py requires GROMACS and MDAnalysis library to be installed.
%\newline \\
%Run AddData.py as follows:
%\newline \\
%python3 AddData.py -f exampleINFO.yaml
%\newline \\
%There is also a script runAddData.sh that can be used to loop over several info files to add many simulations at one go.

%\subsubsection*{Simulation dictionary}
%A simulation dictionary contains information about the simulation. Some of the information is provided by the user. The numbers of lipid molecules, solvent and ions are automatically read from the files and so are the simulation temperature and trajectory length. All this information is saved to a file named README.yaml.

%\subsubsection{User input}
%Dictionary variables defined by the user are listed here. 
%The necessary parameters required for the analyses are described first and are marked as ''compulsory''. Then the parameters used to describe the data for further usage and analyses are described.
%While only compulsory parameters are required for the databank entry, strong recommendation is that all the possible parameters would be given to maximize the possibilities to upcycle the contributed data.

>>>>

\title{SUPPLEMENTARY INFORMATION: NMRlipids Databank: Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration}

@@

\begin{abstract}
%We present a databank of lipid bilayer simulations from the NMRlipids open collaboration project.
\end{abstract}

\begin{document}

@@

\pagebreak
\section{Water permeation through membranes}

\begin{figure}[!h]
    \centering
    \includegraphics[width = 180mm]{Figures/permeationSI.pdf}
    \caption{a) Structure of an overlay databank. 
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    The figures and numbers are created on 9th of May 2022.}
    \label{fig:overlay}
\end{figure}

\pagebreak
\section{Water diffusion along membranes}

\begin{figure}[!h]
    \centering
    \includegraphics[width = 180mm]{Figures/LateralDiffusionSI.pdf}
    \caption{a) Structure of an overlay databank. 
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    The figures and numbers are created on 9th of May 2022.}
    \label{fig:overlay}
\end{figure}
